Variance reduction for MCMC methods via martingale representations

### 2020-02-18 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex

@@

A_{s,k}(x)=\frac{1}{s}\sum_{r=0}^{s}\bar a_{r,k}(x).

@@

h(y_1,\ldots,z_m),

@@

Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K\geq 2$ times continuously differentiable

@@

\begin{lem}\label{eq:a_repr}
Fix some \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) with \(\mathbf{k}'\le \mathbf{k}\) componentwise.  Then the following representation holds
\[
\bar a_{p,\mathbf{k}}(x)=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}') !}{\mathbf{k}!}\right)^{1/2}
\PE\left[
\partial_{1}^{\mathbf{k}'}f(X^x_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_l)\right] \eqsp.
\]
\end{lem}

\begin{proof}
Let $\boldsymbol{\varphi}(z)=\frac{1}{(2\pi)^{d/2}} \exp\{-|z|^2/2\}$,
$z\in\mathbb R^d$, denote the density of a $d$-dimensional
standard Gaussian random vector.
We first remark that, for the normalized Hermite polynomial $\mathbf H_{\mathbf k}$ on $\mathbb R^d$,
$\mathbf k\in\mathbb N_0^d$, it holds
$$
\mathbf{H}_{\mathbf{k}}(z)\boldsymbol{\varphi}(z)
=\frac{(-1)^{|\mathbf{k}|}}{\sqrt{\mathbf{k} !}} \partial^{\mathbf{k}} \boldsymbol{\varphi}(z).
$$
This enables to use the integration by parts in vector form as follows
(below $\prod_{j=l+1}^p:=1$ whenever $l=p$)
\begin{align*}
&\bar a_{p,\mathbf{k}}(x)
 =
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
\mathbf{H}_{\mathbf{k}}(z_{1})\boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
\\
& =
\frac{1}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
(-1)^{|\mathbf{k}|}\partial^{\mathbf{k}} \boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
\\
& =
\frac{\gamma^{|\mathbf k'|/2}}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
\partial_{1}^{\mathbf k'}[f\circ G_{p}](x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
(-1)^{|\mathbf{k}-\mathbf k'|}\partial^{\mathbf{k}-\mathbf k'} \boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, dz_{1}\ldots dz_{p}
\\
 & =\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[\partial_{y_{1}}^{\mathbf{k}'}[f\circ G_{p}](x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p})\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right].
\end{align*}
The last expression yields the result.
\end{proof}

For multi-indices $\mathbf k,\mathbf k'\in\mathbb N_0^d$
with $\mathbf k'\le\mathbf k$ componentwise
and $\mathbf k'\ne\mathbf k$,
we get applying first Lemma~\ref{eq:a_repr}
\begin{equation*}
A_{s,\mathbf{k}}(x)
=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
\,
\mathsf{E}_x\left[
\frac{1}{s}\sum_{r=1}^{s}(\partial_{1}^{\mathbf{k}'} f(X_r)-\mathsf{E}_x\partial_{1}^{\mathbf{k}'}f(X_r))\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
\end{equation*}

Assume that $\mu$ and $f$ are $K\times d$ times continuously differentiable.
Then, given $\mathbf k\in\mathbb N_0^d$,
by taking $\mathbf k'= \mathbf k'(\mathbf k)
=(K1_{\{k_1>K\}},\ldots,K1_{\{k_d>K\}})$, we get
\begin{align}
\label{eq:sum_abar}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x) & =\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q_s(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
\nonumber
 & =\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}
 \\
 \nonumber
 & \hspace{1em}\times\sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q_s(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}),
\end{align}
where for any two multi-indices \(\mathbf{r},\) \(\mathbf{q}\) from \(\mathbb{N}_0^d\)
\begin{eqnarray*}
Q_s(\mathbf{r},\mathbf{q})
=
\left(\mathsf{E}_x\left[\frac{1}{s}\sum_{p=1}^{s}\left(\partial_{y_{1}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}_x\left[\partial_{y_{1}}^{\mathbf r}f\left(X_{p}\right)\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right)^{2}.
\end{eqnarray*}
In \eqref{eq:sum_abar} the first sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}.$
For any subset $I,$ $\mathbb{N}_{I}^{d}$ stands for a set
of multi-indices $\mathbf{m}_{I}$ with elements $m_{i}=0,$ $i\not\in I,$
and $m_{i}\in\mathbb{N},$  $i\in I.$ Moreover, \(I^c=\{1,\ldots,d\}\setminus I\) and \(\mathbb{N}^d_{0,I^c}\) stands for a set
of multi-indices $\mathbf{m}_{I^c}$ with elements $m_{i}=0,$ $i\in I,$
and $m_{i}\in\mathbb{N}_0,$  $i\not\in I.$ Finally, the multi-index \(\mathbf{K}_I\) is defined as $\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).$
Applying the estimate
\begin{eqnarray*}
\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}\leq (1/2)^{|I| K},
\end{eqnarray*}
we get
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)&\leq &
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K}
\\
&& \times\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}} \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c})
\\
&\leq &
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K} \sum_{\mathbf{m}\in\mathbb{N}_0^{d}} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}).
\end{eqnarray*}
Now using the consequence
$\sum_{\mathbf{m}\in \mathbb{N}^d_0} \langle\xi,\mathbf{H}_\mathbf{m}(Z_1)\rangle^2\le\langle\xi,\xi\rangle$
of the Parseval's identity,
we derive
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
&\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X_{p}\right)
\right)
\end{eqnarray*}
Next we show that under the conditions  of Theorem~\ref{th:mr}
\begin{eqnarray*}
\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X_{p}\right)
\right)\leq W_s(x)
\end{eqnarray*}
for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$

>>>>

\label{eq:definition-bar-a}

@@

\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).

@@

h(y_1,\ldots,y_m),

@@

Suppose additionally that a bounded function $f$ and $\mu=\nabla U$ are $K \times d \geq 2$ times continuously differentiable

@@

\begin{lem}\label{eq:a_repr}
Fix some \(\mathbf{k},\mathbf{k}'\in \mathbb{N}_0^d\) with \(\mathbf{k}'\le \mathbf{k}\) componentwise.  Then the following representation holds
\[
\bar a_{p,\mathbf{k}}(x)=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}') !}{\mathbf{k}!}\right)^{1/2}
\PE\left[
\partial_{1}^{\mathbf{k}'}f(X^x_p)\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right] \eqsp,
\]
where $\bar a_{p,\mathbf{k}}$ is defined in \eqref{eq:definition-bar-a}.
\end{lem}

\begin{proof}
Let $\boldsymbol{\varphi}(z)=(2\pi)^{-d/2} \exp(-|z|^2/2)$,
$z\in\mathbb R^d$, denote the density of a $d$-dimensional
standard Gaussian random vector.
We first remark that, for the normalized Hermite polynomial $\mathbf H_{\mathbf k}$ on $\mathbb R^d$,
$\mathbf k\in\mathbb N_0^d$, it holds
$$
\mathbf{H}_{\mathbf{k}}(z)\boldsymbol{\varphi}(z)
=\frac{(-1)^{|\mathbf{k}|}}{\sqrt{\mathbf{k} !}} \partial^{\mathbf{k}} \boldsymbol{\varphi}(z).
$$
This enables to use the integration by parts in vector form as follows
(below $\prod_{j=l+1}^p:=1$ whenever $l=p$)
\begin{align*}
&\bar a_{p,\mathbf{k}}(x)
 =
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
\mathbf{H}_{\mathbf{k}}(z_{1})\boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots \rmd z_{p}
\\
& =
\frac{1}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
f\circ G_{p}(x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
(-1)^{|\mathbf{k}|}\partial^{\mathbf{k}} \boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots  \rmd z_{p}
\\
& =
\frac{\gamma^{|\mathbf k'|/2}}{\sqrt{\mathbf k!}}
\int_{\mathbb R^d}\ldots\int_{\mathbb R^d}
\partial_{1}^{\mathbf k'}[f\circ G_{p}](x,\sqrt{\gamma}z_{1},\ldots,\sqrt{\gamma}z_{p})
(-1)^{|\mathbf{k}-\mathbf k'|}\partial^{\mathbf{k}-\mathbf k'} \boldsymbol{\varphi}(z_1)
\prod_{j=2}^p\boldsymbol{\varphi}(z_j)\, \rmd z_{1}\ldots \rmd z_{p}
\\
 & =\gamma^{|\mathbf{k}'|/2}\frac{\sqrt{(\mathbf{k}-\mathbf{k}')!}}{\sqrt{\mathbf{k}!}}\mathsf{E}\left[\partial_{y_{1}}^{\mathbf{k}'}[f\circ G_{p}](x,\sqrt{\gamma}Z_{1},\ldots,\sqrt{\gamma}Z_{p})\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_{1})\right].
\end{align*}
The last expression yields the result.
\end{proof}

For multi-indices $\mathbf k,\mathbf k'\in\mathbb N_0^d$
with $\mathbf k'\le\mathbf k$ componentwise
and $\mathbf k'\ne\mathbf k$,
we get applying first Lemma~\ref{eq:a_repr},
\begin{equation*}
A_{s,\mathbf{k}}(x)
=\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)^{1/2}
\,
\PE\left[
s^{-1} \sum\nolimits_{r=1}^{s}\{\partial_{1}^{\mathbf{k}'} f(X^x_r)-\PE[\partial_{1}^{\mathbf{k}'}f(X^x_r)]\}\mathbf{H}_{\mathbf{k}-\mathbf{k}'}(Z_1)\right]
\end{equation*}
where $A_{s,\mathbf{k}}$ is defined in \eqref{eq:definition-A-s,k}.
Assume that $\mu$ and $f$ are $K\times d$ times continuously differentiable.
Then, given $\mathbf k\in\mathbb N_0^d$,
by taking $\mathbf k'= \mathbf k'(\mathbf k)
=(K1_{\{k_1>K\}},\ldots,K1_{\{k_d>K\}})$, we get
\begin{align}
\label{eq:sum_abar}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x) & =\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}\left(\gamma^{|\mathbf{k}'|}\frac{(\mathbf{k}-\mathbf{k}')!}{\mathbf{k}!}\right)Q_s(\mathbf{k}',\mathbf{k}-\mathbf{k}')\\
\nonumber
 & =\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}\gamma^{|I|K}\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}}\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}
 \\
 \nonumber
 & \hspace{1em}\times\sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K}Q_s(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c}),
\end{align}
where for any two multi-indices \(\mathbf{r},\) \(\mathbf{q}\) from \(\mathbb{N}_0^d\)
\begin{eqnarray*}
Q_s(\mathbf{r},\mathbf{q})
=
\left(\mathsf{E}_x\left[\frac{1}{s}\sum_{p=1}^{s}\left(\partial_{y_{1}}^{\mathbf r}f\left(X_{p}\right)-\mathsf{E}_x\left[\partial_{y_{1}}^{\mathbf r}f\left(X_{p}\right)\right]\right)\mathbf{H}_{\mathbf{q}}(Z_{1})\right]\right)^{2}.
\end{eqnarray*}
In \eqref{eq:sum_abar} the first sum runs over all nonempty subsets $I$ of the set $\{1,\ldots,d\}.$
For any subset $I,$ $\mathbb{N}_{I}^{d}$ stands for a set
of multi-indices $\mathbf{m}_{I}$ with elements $m_{i}=0,$ $i\not\in I,$
and $m_{i}\in\mathbb{N},$  $i\in I.$ Moreover, \(I^c=\{1,\ldots,d\}\setminus I\) and \(\mathbb{N}^d_{0,I^c}\) stands for a set
of multi-indices $\mathbf{m}_{I^c}$ with elements $m_{i}=0,$ $i\in I,$
and $m_{i}\in\mathbb{N}_0,$  $i\not\in I.$ Finally, the multi-index \(\mathbf{K}_I\) is defined as $\mathbf{\mathbf{K}}_{I}=(K1_{\{1\in I\}},\ldots,K1_{\{d\in I\}}).$
Applying the estimate
\begin{eqnarray*}
\frac{\mathbf{m}_{I}!}{\left(\mathbf{m}_{I}+\mathbf{K}_{I}\right)!}\leq (1/2)^{|I| K},
\end{eqnarray*}
we get
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)&\leq &
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K}
\\
&& \times\sum_{\mathbf{m}_{I}\in\mathbb{N}_{I}^{d}} \sum_{\mathbf{m}_{I^c}\in \mathbb{N}^d_{0,I^c},\,\|\mathbf{m}_{I^c}\|\leq K} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}_{I}+\mathbf{m}_{I^c})
\\
&\leq &
\sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset} (\gamma_{l}/2)^{|I|K} \sum_{\mathbf{m}\in\mathbb{N}_0^{d}} Q(\mathbf{\mathbf{K}}_{I},\mathbf{m}).
\end{eqnarray*}
Now using the consequence
$\sum_{\mathbf{m}\in \mathbb{N}^d_0} \langle\xi,\mathbf{H}_\mathbf{m}(Z_1)\rangle^2\le\langle\xi,\xi\rangle$
of the Parseval's identity,
we derive
\begin{eqnarray*}
\sum_{\mathbf{k}\colon\|\mathbf{k}\|\geq K+1}A^2_{s,\mathbf{k}}(x)
&\leq & \sum_{I\subseteq\{1,\ldots,d\},\, I\neq \emptyset}
\left(\frac{\gamma_{l}}{2}\right)^{|I|K}
\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X_{p}\right)
\right)
\end{eqnarray*}
Next we show that under the conditions  of Theorem~\ref{th:mr}
\begin{eqnarray*}
\mathsf{Var}_x\left(\frac{1}{s}\sum_{p=1}^{s}\partial_{1}^{\mathbf{K}_I}f\left(X_{p}\right)
\right)\leq W_s(x)
\end{eqnarray*}
for all $x$ and some family of functions $(W_s)_{s\geq 1}$ such that  the sum $\sum_{s=1}^n \mathsf{E}[W_s(X_{N+n-s-1})]$ is bounded uniformly in $n\in \mathbb{N}.$
